---
title: "Example of use of the functions for parsing pubmed records"
output: html_document
author: "juarpasi"
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

# Introduction
This vignette show how to use the `getMetadata` function in one PubMed record.

# Preparation

If you have a _entrez key_ you can save it in the variable `ENTREZ_KEY`.
For this example I will use the `easyPubMed` package to get one article.
```{r, warning = F, message = F}
# put below your own entrez key if you have one
ENTREZ_KEY=NULL
library(easyPubMed)
library(xml2)

#copy the funcitons in your working directory and export them.
#if you have the functions in other place, you need to specify the path
source('getMetadata.R')
source('getAuthors.R')
source('makeDfFromXmlTxt.R')
source('nodeList2Df.R')
```
To make the functions work, we need to use the `xml2` package.

Next, we need at least one PubMed record.
```{r, warning = F, message = F}
query <- 'cancer[TI] AND "2020"[dp] AND "journal article"[pt] AND "free full text"[sb] AND adolescent[mh]'

search <- get_pubmed_ids(query, api_key = ENTREZ_KEY)

oneRecordCancer2020 <- fetch_pubmed_data(search, retmax = 1)

doc <- xml2::read_xml(oneRecordCancer2020)|>
  xml2::xml_find_all(".//PubmedArticle")
```

The `oneRecordCancer2020` object contains one PubMed record.
This record will be used for parsing.
But first, we need to convert it into a `xml2` document since the `getMetadata` function work with this object.

# Parse the PubMed record
To extract the abstract, we place it in the `metadata.list` argument.
```{r, warning = F, message = F}
recordAbs <-
  getMetadata(doc, metadata.list = 'abstract')
```

```{r,warning = F, message = F}
datatable(recordAbs[["abstract"]])
```

You can specify other metadata fields if you are interested in them.
```{r,warning = F, message = F}
recordMetadata <-
  getMetadata(doc,
              metadata.list = c('keywordTerms','journal'))
```

```{r,warning = F, message = F}
datatable(recordMetadata$keywordTerms)
datatable(recordMetadata$journal)
```

Currently, the available metadata fields that can be parsed in the `metadata.list` argument are the follows:

* abstract
* meshDescrip
* meshQualifier
* keywordTerms
* pubType
* journal
* country
* language
* title
* year
* authors

Due to the structure of the PubMed record, each metadata field requires one of the following functions `getAuthors.R`, `makeDfFromXmlTxt` or `nodeList2Df`.

```{r, echo=FALSE}
sessionInfo()
```