---
title: "How to parsed multiple PubMed records"
output: html_document
author: "juarpasi"
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

# Introduction
The `getMetadata` function works well with one PubMed record.
However, there are cases where we need to parse multiple records

# Preparation

If you have a _entrez key_ you can save it in the variable `ENTREZ_KEY`.
For this example I will use the `easyPubMed` package to retrieve the articles.
```{r, warning = F, message = F}
# put below your own entrez key if you have one
ENTREZ_KEY=NULL
library(easyPubMed)
library(xml2)
library(dplyr)
```
To make the functions work, we need to use the `xml2` package.

# Get pubmed records
Next, we will download approximately 1500 PubMed record.
```{r}
query <- '"breast neoplasms, male"[MeSH Terms] AND "journal article"[pt] AND ("2010"[dp]:"2022"[dp])'

#download the records in the working directory
batch_pubmed_download(query,
                      #dest_dir = 'data',
                      dest_file_prefix = paste0("maleCancer_"),
                      format = "xml", api_key = ENTREZ_KEY,
                      batch_size = 400)
```

In this case, the records are saved in `.txt` files.
We need to read them into the `R` environment.
Additionally, we also need to use others packages to parallelize the work.

```{r, warning = F, message = F}
library(XML)
library(xml2)
library(foreach)
library(doParallel)
library(ggplot2)

#copy the funcitons in your working directory and export them.
#if you have the function in other place, you need to specify the path
source('getMetadata.R')
#a simple function for read the files with pubmeds records
source('readXML.R')
```

# Parse metadata
```{r, warning = F, message = F}
files <- dir(pattern = 'maleCancer*')
files <- lapply(files, readXMLPubmed)|> Reduce(c, x= _)

###########################################
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

# We use the foreach pkg to iterate on the nodeset and call the function
metadata <- foreach(file = files, .packages = c("xml2")) %dopar% {
  
  source('getAuthors.R')
  source('makeDfFromXmlTxt.R')
  source('nodeList2Df.R')

  doc <- xml2::read_xml(file)
  r <- getMetadata(doc,
                   metadata.list = c('meshDescrip','meshQualifier','journal','year'))
  return(r)
}

stopCluster(cl)
###########################################

#clean memory
rm(cl)
rm(files)
gc()
```

# Make some analysis
Now, the `metadata` object is a list that contains the metadata for each record obtained.
We can use the `data.table` package to get a dataframe for each data.

We can start to make some analysis on the metadata.
For example, we can create a histogram showing the publication years.
```{r, warning = F, message = F}
years <- lapply(metadata, `[[`, 'year') |> data.table::rbindlist(fill=TRUE)

papersYearsPlot <-
  years |>
  as_tibble() |>
  count(Year) |>
  arrange() |>
  mutate(across(Year, as.integer)) |>
  ggplot() + 
  geom_col(aes(x = Year, y = n), fill = '#4271AE')+
  labs(x = 'year',
       y = 'N of arcticles',
       title = 'Papers of male cancer') +
  theme(plot.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill=NA, size = 0.8),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.text.x = element_text(angle = -90, vjust = 0.5))

papersYearsPlot
```

Which are the main journals that publish related to the query?
```{r, warning = F, message = F}
library(tidyr)
journals <- lapply(metadata, `[[`, 'journal') |> data.table::rbindlist(fill=TRUE)

journalsTbl <-
  journals |>
  select(Title) |>
  drop_na() |>
  count(Title) |>
  slice_max(n, n = 20) |>
  arrange(desc(n))

datatable(journalsTbl)
```

What are the most frequent MeSH terms in the records?
```{r, warning = F, message = F}
library(forcats)

meshD <- lapply(metadata, `[[`, 'meshDescrip') |> data.table::rbindlist(fill=TRUE)
meshQ <- lapply(metadata, `[[`, 'meshDescrip') |> data.table::rbindlist(fill=TRUE)

meshD$Type <- NULL
meshD <-
  meshD |>
  select(id, text) |>
  mutate(type = 'descriptor')

meshQ <-
  meshQ |>
  select(id, text) |>
  mutate(type = 'qualifiers')

mesh <-
  data.table::rbindlist(list(meshD,meshQ))

meshFreqPlot <-
  mesh |>
  drop_na() |>
  count(text, type) |>
  arrange(desc(n))|>
  slice_max(order_by = n, n = 25) |>
  mutate(meshes = fct_reorder(text, n)) |>
  ggplot() +
  geom_col(aes(x = n, y = meshes, fill = type)#,fill = 'red'
           )+
  labs(x = 'Mesh Terms',
       y = '',
       title = 'Most common mesh terms') +
  theme(plot.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill=NA, size = 0.8),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.text.x = element_text(angle = -90, vjust = 0.5))

meshFreqPlot
```


```{r, echo=FALSE}
sessionInfo()
```