---
title: "How to parse multiple PubMed records"
output: 
  html_document:
    toc: true
    toc_depth: 2
author: "juarpasi"
date: "2023-05-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

## Introduction
The `getMetadata` function works well with one PubMed record.
However, there are cases where we need to parse multiple records.

This tutorial shows how can we extract the metadata from multiple PubMed records retrieved with the `easyPubMed` package.

# Parsing PubMed records

## Preparation

If you have a _entrez key_ you can save it in the variable `ENTREZ_KEY`.
```{r, warning = F, message = F}
# put below your own entrez key if you have one
ENTREZ_KEY=NULL
library(easyPubMed)
library(xml2)
library(dplyr)
```

## Get pubmed records
Next, we will download approximately 1500 PubMed record.
```{r}
query <- '"breast neoplasms, male"[MeSH Terms] AND "journal article"[pt] AND ("2010"[dp]:"2022"[dp])'

#download the records in the working directory
batch_pubmed_download(query,
                      dest_file_prefix = "maleCancer_",
                      format = "xml", api_key = ENTREZ_KEY,
                      batch_size = 400)
```

In this case, the records are saved in `.txt` files.
We need to read them into the `R` environment.
The `readXMLPubmed` function in the `readXML.R` _script_ read one file containing the PubMeds Records.

Additionally, we also need to use others packages to parallelize the work.

```{r, warning = F, message = F}
library(XML)
library(foreach)
library(doParallel)

#copy the funcitons in your working directory and export them.
#if you have the function in other place, you need to specify the path
source('getMetadata.R')
source('readXML.R')
```

## Parse metadata
```{r, warning = F, message = F, results='hide'}
files <- dir(pattern = 'maleCancer*')
files <- lapply(files, readXMLPubmed)|> Reduce(c, x= _)

###########################################
cl <- makeCluster(detectCores() - 1)
registerDoParallel(cl)

# We use the foreach pkg to iterate on the nodeset and call the function
metadata <- foreach(file = files, .packages = c("xml2")) %dopar% {
  
  #call the functions
  source('getAuthors.R')
  source('makeDfFromXmlTxt.R')
  source('nodeList2Df.R')

  dat <-
    xml2::read_xml(file) |>
    getMetadata(
      #add the metada you want to obtain
      metadata.list = c('meshDescrip','meshQualifier','journal','year'))
  
  return(dat)
}

stopCluster(cl)
###########################################

#clean memory
rm(cl)
rm(files)
gc()
```

# Metada Anlysis

## Convert list into a dataframe
Now, the `metadata` object is a list that contains the metadata for each record obtained.
We can use the `data.table` package to get a dataframe for each data.

## Paper per year
We can start to make some analysis on the metadata.
For example, we can create a histogram showing the publication years.
```{r, warning = F, message = F}
library(ggplot2)
years <- lapply(metadata, `[[`, 'year') |> data.table::rbindlist(fill=TRUE)

papersYearsPlot <-
  years |>
  as_tibble() |>
  count(Year) |>
  arrange() |>
  mutate(across(Year, as.integer)) |>
  ggplot() + 
  geom_col(aes(x = Year, y = n), fill = '#4271AE')+
  labs(x = 'year',
       y = 'N of arcticles',
       title = 'Papers of male cancer') +
  theme(plot.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill=NA, size = 0.8),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.text.x = element_text(angle = -90, vjust = 0.5))

papersYearsPlot
```

## Journals
Which are the main journals that publish related to the query?
```{r, warning = F, message = F}
library(tidyr)
journals <- lapply(metadata, `[[`, 'journal') |> data.table::rbindlist(fill=TRUE)

journalsTbl <-
  journals |>
  select(Title) |>
  drop_na() |>
  count(Title) |>
  slice_max(n, n = 20) |>
  arrange(desc(n))

datatable(journalsTbl)
```

## Analyzing MeSH terms
What are the most frequent MeSH terms in the records?
```{r, warning = F, message = F}
library(forcats)

meshD <- lapply(metadata, `[[`, 'meshDescrip') |> data.table::rbindlist(fill=TRUE)
meshQ <- lapply(metadata, `[[`, 'meshDescrip') |> data.table::rbindlist(fill=TRUE)

meshD$Type <- NULL
meshD <-
  meshD |>
  select(id, text) |>
  mutate(type = 'descriptor')

meshQ <-
  meshQ |>
  select(id, text) |>
  mutate(type = 'qualifiers')

mesh <-
  data.table::rbindlist(list(meshD,meshQ))

meshFreqPlot <-
  mesh |>
  drop_na() |>
  count(text, type) |>
  arrange(desc(n))|>
  slice_max(order_by = n, n = 25) |>
  mutate(meshes = fct_reorder(text, n)) |>
  ggplot() +
  geom_col(aes(x = n, y = meshes, fill = type))+
  labs(x = 'Mesh Terms',
       y = '',
       title = 'Most frequent MeSH terms') +
  theme(plot.title = element_text(face = "bold"),
        panel.border = element_rect(color = "black", fill=NA, size = 0.8),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = "transparent",colour = NA),
        plot.background = element_rect(fill = "transparent",colour = NA),
        axis.text.x = element_text(angle = -90, vjust = 0.5))

meshFreqPlot
```


```{r, echo=FALSE}
sessionInfo()
```